Hydra Battles and Cie

### 2021-02-09 doc/chapter-powers.tex
proof simplification, typos in doc

\subsection{A common notation for multiplication}
\label{op-classes}
\index{Coq!Type Classes!Operational Type Classes}
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

\subsection{The Monoid Type Class}
\index{Coq!Type Classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\index{Coq!Type Classes}
\index{Coq!Proper}
\label{EMonoid-def}

@@

In order to make possible proof by rewriting on expressions which contain
the exponentiation operator, we have to prove that, whenever \texttt{$x$ == $y$},
the equality \texttt{$x^n$ == $y^n$} holds for any exponent \texttt{$n$}. 
For this purpose, we use the \texttt{Proper} class of module
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.Morphisms.html}{Coq.Classes.Morphisms}
\index{Coq!Type Classes}
\index{Coq!Proper}
\begin{Coqsrc}
Global Instance power_proper :
     Proper (equiv ==> eq ==> equiv) power.
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{itemize}
\item Continuation Passing Style (CPS) \index{Programming Styles!Continuation Passing Style (CPS)} \cite{reynolds93}
is a way to make explicit the 
     control in the evaluation of an expression, in a purely functional way. 
    For every intermediate computation step, the result is sent
    to a \emph{continuation} that executes the further continuations.
   When the continuation is a lambda-abstraction, its bound variable 
   gives a \emph{name} to this result

@@

@@

\subsubsection{Definition:} 
\index{Coq!Type Classes}
\index{Coq!Proper}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

\index{Programming Styles!Continuation Passing Style (CPS)}

@@

The following dependently-typed functions will help us to specify  formally
any correct chain generator.
\index{Coq!Dependently Typed Functions}

>>>>

\subsection{A common notation for multiplication}
\label{op-classes}
\index{Coq!Type classes!Operational type classes}
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

\subsection{The Monoid Type Class}
\index{Coq!Type classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\index{Coq!Type classes}
\index{Coq!Proper}
\label{EMonoid-def}

@@

In order to make possible proof by rewriting on expressions which contain
the exponentiation operator, we have to prove that, whenever \texttt{$x$ == $y$},
the equality \texttt{$x^n$ == $y^n$} holds for any exponent \texttt{$n$}. 
For this purpose, we use the \texttt{Proper} class of module
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.Morphisms.html}{Coq.Classes.Morphisms}
\index{Coq!Type classes}
\index{Coq!Proper}
\begin{Coqsrc}
Global Instance power_proper :
     Proper (equiv ==> eq ==> equiv) power.
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{itemize}
\item Continuation Passing Style (CPS) \index{Programming styles!Continuation Passing Style (CPS)} \cite{reynolds93}
is a way to make explicit the 
     control in the evaluation of an expression, in a purely functional way. 
    For every intermediate computation step, the result is sent
    to a \emph{continuation} that executes the further continuations.
   When the continuation is a lambda-abstraction, its bound variable 
   gives a \emph{name} to this result

@@

\index{Coq!Plug-ins!paramcoq}

@@

\subsubsection{Definition:} 
\index{Coq!Type classes}
\index{Coq!Proper}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

\index{Programming styles!Continuation Passing Style (CPS)}

@@

The following dependently-typed functions will help us to specify  formally
any correct chain generator.
\index{Coq!Dependently typed functions}

### 2021-02-09 doc/chapter-primrec.tex
proof simplification, typos in doc

This chapter contains some comments on Russel's library, as well as a few extensions.
Contributions (under the form of comments, new examples or exercises) are welcomed!.

\section{First look at the Ackermann library}

@@

\section{Basic Libraries}

@@

The library allows us to consider primitive functions on \texttt{nat}, with any number of arguments, in 
curried form. This is made possible in 
\href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat} by the following definition:

@@

\subsection{A little bit of Semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ (and the helper function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We leave it to the reader  to look also at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

 with evalPrimRecs (n m : nat) (fs : PrimRecs n m) {struct fs} :
 Vector.t (naryFunc n) m :=
  match fs in (PrimRecs n m) return (Vector.t (naryFunc n) m) with
  | PRnil a => Vector.nil  (naryFunc a)
  | PRcons a b g gs =>
       Vector.cons _ (evalPrimRec _ g) _  (evalPrimRecs _ _ gs)
  end.
\end{Coqsrc}
Let us take as example the following term\footnote{Of course, we never typed this term \emph{verbatim}; we obtained it as the result of some computation we leave it the reader to guess and reproduce.}:

\label{sect:weirdfac}
\begin{Coqsrc}
Example weirdPR : PrimRec 1 :=
primRecFunc 0
  (composeFunc 0 1 (PRcons 0 0 zeroFunc (PRnil 0)) succFunc)
  (composeFunc 2 2
    (PRcons 2 1
      (composeFunc 2 1
         (PRcons 2 0 (projFunc 2 1 (le_n 2))
                 (PRnil 2))
         succFunc)
      (PRcons 2 0
        (composeFunc 2 1
          (PRcons 2 0
             (projFunc 2 0
                       (le_S 1 1 (le_n 1)))
             (PRnil 2))
          (projFunc 1 0 (le_n 1))) (PRnil 2)))
    (primRecFunc 1 (composeFunc 1 0 (PRnil 1) zeroFunc)
       (composeFunc 3 2
         (PRcons 3 1
            (projFunc 3 1 (le_S 2 2 (le_n 2)))
            (PRcons 3 0 (projFunc 3 0
                          (le_S 1 2
                                (le_S 1 1 (le_n 1))))
                    (PRnil 3)))
         (primRecFunc 1 (projFunc 1 0 (le_n 1))
                      (composeFunc 3 1
                          (PRcons 3 0
                                  (projFunc 3 1 (le_S 2 2 (le_n 2)))
                                  (PRnil 3))
                          succFunc))))). 
\end{Coqsrc}

@@

\begin{Coqsrc}
Example  mystery_fun : nat _> nat := evalPrimRec 1 weirdPR.

Compute List.map mystery_fun (0::1::2::3::4::5::6::nil) : list nat.
\end{Coqsrc}

\begin{Coqanswer}
 = 1 :: 1 :: 2 :: 6 :: 24 :: 120 :: 720 :: nil
     : list nat
\end{Coqanswer}

Ok, the term \texttt{weirdPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections much simpler ways to prove that a given function is primitive recursive, whithout looking at an unreadable term.

@@

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}. Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. Our example of page\pageref{sect:weirdfac} was built using a  copy of \texttt{primRec.v} where many \texttt{Qed}s have been replaced with
\texttt{Defined}s.
\end{remark}

@@

The best known of total function which is not primitive recursive is the Ackermann function. We show how to adapt the classic proof (see for instance~\cite{planetmath}) to the constraints of \gallina.

@@

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.MoreAck.Ack.html}{Ackermann.MoreAck.Ack}}.

@@

\begin{Coqsrc}
Lemma Ack_1_n n : Ack 1 n = S (S n).

@@

Please note that for any given $n$, the unary function
(\texttt{Ack\,$n$} is primitive recursive.

@@

\begin{enumerate}
\item Prove that for any primitive recursive function $f:\mathbb{N}\rightarrow\mathbb{N}\rightarrow\mathbb{N}$, there exists some natural number $n$ depending on $f$, such that, for any $x$ and $y$, 
$f\,x\,y \leq \texttt{Ack}\,n\,(\textrm{max}\,x\,y)$ (we say that $f$ is \emph{majorized}  by \texttt{Ack}).
\item Show that \texttt{Ack} fails to satisfy this property.
\end{enumerate}

@@

We are now able to translate in \gallina{} the notion of majoration:

@@

\begin{Coqsrc}
 ...
-  (** empty list of functions *)
\end{Coqsrc}

@@

  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn :  forall x y,  Ack x y <= Ack n (max x  y) .

    Remark R01 : n <> 0.

@@

    Remark R03 : max 2 n = n.

    Remark R04 : Ack n (2 * n + 3) = Ack n (Ack 2 n).

    Remark R05 : Ack n (2 * n + 3) <= Ack (n + 2) n.

    Remark R06 : Ack (n + 2) n <= Ack n (2 * n + 2).

    Remark R07 : Ack n (2 * n + 3) <= Ack n (2 * n + 2).

    Remark R08 :  Ack n (2 * n + 2) < Ack n (2 * n + 3).

>>>>

This chapter contains some comments on Russel's library, as well as a few extensions.
Contributions (under the form of comments, new examples or exercises) are welcome!.

\section{First look at the Ackermann Library}

@@

\section{Basic Definitions}
\index{Maths!Primitive recursive functions}

@@

Library \texttt{primRec} allows us to consider primitive functions on \texttt{nat}, with any number of arguments, in 
curried form. This is made possible in 
\href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat} by the following definition:

@@

\subsection{A little bit of Semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ , together with the helper function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

\index{Coq!Dependent pattern matching}
Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We leave it to the reader  to look also at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

 with evalPrimRecs (n m : nat) (fs : PrimRecs n m) {struct fs} :
 Vector.t (naryFunc n) m :=
  match fs in (PrimRecs n m) return (Vector.t (naryFunc n) m) with
  | PRnil a => Vector.nil  (naryFunc a)
  | PRcons a b g gs =>
       Vector.cons _ (evalPrimRec _ g) _  (evalPrimRecs _ _ gs)
  end.
\end{Coqsrc}

Looks complicated? The following examples show that, when
the arity is fixed, these definitions behave well w.r.t. 
\coq's reduction rules. Moreover, they make the interpretation functions more ``concrete''.

\begin{Coqsrc}
Example Ex1 : evalPrimRec 0 zeroFunc = 0.
Proof. reflexivity. Qed.

Example Ex2 a : evalPrimRec 1 succFunc a = S a.
Proof. reflexivity. Qed.

Example Ex3 a b c d e f: forall (H: 2 < 6),
    evalPrimRec 6
                (projFunc 6 2 H) a b c d e f = d.
Proof. reflexivity. Qed.

Example Ex4 (x y z : PrimRec 2) (t: PrimRec 3):
  let u := composeFunc 2 3
                       (PRcons 2 2 x
                               (PRcons 2 1 y
                                       (PRcons 2 0 z
                                               (PRnil 2))))
                       t in
  let f := evalPrimRec 2 x in
  let g := evalPrimRec 2 y in
  let h := evalPrimRec 2 z in
  let i := evalPrimRec 3 t in
  let j := evalPrimRec 2 u in
  forall a b, j a b = i (f a b) (g a b) (h a b).
Proof. reflexivity. Qed.

Example Ex5 (x : PrimRec 2)(y: PrimRec 4):
  let g := evalPrimRec _ x in
  let h := evalPrimRec _ y in
  let f := evalPrimRec _ (primRecFunc _ x y) in
  forall a b,  f 0 a b = g a b.
Proof. reflexivity.   Qed.

Example Ex6 (x : PrimRec 2)(y: PrimRec 4):
  let g := evalPrimRec _ x in
  let h := evalPrimRec _ y in
  let f := evalPrimRec _ (primRecFunc _ x y) in
  forall n a b,  f (S n) a b = h n (f n a b) a b.
Proof. reflexivity.   Qed.                          
\end{Coqsrc}

Another example?
Let us consider the following term\footnote{Of course, we never typed this term \emph{verbatim}; we obtained it as the result of some computation we leave it the reader to guess and reproduce.}:

\label{sect:bigfac}
\begin{Coqsrc}
Example bigPR : PrimRec 1 :=
primRecFunc 0
  (composeFunc 0 1 (PRcons 0 0 zeroFunc (PRnil 0)) succFunc)
  (composeFunc 2 2
    (PRcons 2 1
      (composeFunc 2 1
         (PRcons 2 0 (projFunc 2 1 (le_n 2))
                 (PRnil 2))
         succFunc)
      (PRcons 2 0
        (composeFunc 2 1
          (PRcons 2 0
             (projFunc 2 0
                       (le_S 1 1 (le_n 1)))
             (PRnil 2))
          (projFunc 1 0 (le_n 1))) (PRnil 2)))
    (primRecFunc 1 (composeFunc 1 0 (PRnil 1) zeroFunc)
       (composeFunc 3 2
         (PRcons 3 1
            (projFunc 3 1 (le_S 2 2 (le_n 2)))
            (PRcons 3 0 (projFunc 3 0
                          (le_S 1 2
                                (le_S 1 1 (le_n 1))))
                    (PRnil 3)))
         (primRecFunc 1 (projFunc 1 0 (le_n 1))
                      (composeFunc 3 1
                          (PRcons 3 0
                                  (projFunc 3 1 (le_S 2 2 (le_n 2)))
                                  (PRnil 3))
                          succFunc))))). 
\end{Coqsrc}

@@

\begin{Coqsrc}
Example  mystery_fun : nat -> nat := evalPrimRec 1 bigPR.

Compute map mystery_fun[0;1;2;3;4;5;6] : t nat _.
\end{Coqsrc}

\begin{Coqanswer}
 = 1 :: 1 :: 2 :: 6 :: 24 :: 120 :: 720 :: nil
     : t nat 7
\end{Coqanswer}

Ok, the term \texttt{bigPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections much simpler ways to prove that a given function is primitive recursive, whithout looking at an unreadable term.

@@

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}. Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. Our example of page\pageref{sect:bigfac} was built using a  copy of \texttt{primRec.v} where many \texttt{Qed}s have been replaced with
\texttt{Defined}s.
\end{remark}

@@

The best known example of a total recursive function which is not primitive recursive is the Ackermann function. We show how to adapt the classic proof (see for instance~\cite{planetmath}) to the constraints of \gallina. We hope this formal proof 
 is a nice opportunity to explore
the treatment of primitive recursive functions by R. O'Connor,
and to play with dependent types.

@@

Here is a possible definition, where (\texttt{iterate\,$f$\,$n$})
is the $n$-th iteration of $f$.

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Prelude.Iterates.html}{Prelude.Iterates}}.

\begin{Coqsrc}
Fixpoint iterate {A:Type}(f : A -> A) (n: nat)(x:A) :=
  match n with
  | 0 => x
  | S p => f (iterate  f p x)
  end.
\end{Coqsrc}

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.MoreAck.Ack.html}{Ackermann.MoreAck.Ack}}.
\index{Maths!Ackermann function}

@@

\begin{Coqsrc}
Lemma Ack_1_n n: Ack 1 n = S  (S n).

@@

Please note that for any given $n$, the unary function
(\texttt{Ack\,$n$}) is primitive recursive.

@@

\begin{enumerate}
\item Prove that for any primitive recursive function $f:\mathbb{N}\rightarrow\mathbb{N}\rightarrow\mathbb{N}$, there exists some natural number $n$ depending on $f$, such that, for any $x$ and $y$, 
$f\,x\,y \leq \texttt{Ack}\,n\,(\textrm{max}\,x\,y)$ (we say that $f$ is \emph{``majorized''}  by \texttt{Ack}).
\item Show that \texttt{Ack} fails to satisfy this property.
\end{enumerate}

@@

We are now able to translate in \gallina{} the notion of ``majorization'':

@@

\begin{Coqsrc}
 assert (L1 : forall i (v: t nat n) ,
               v_apply f (i::v)  <= Ack q (i + max_v v)).
    { induction i.
      ...
    }
    ...
-  (** empty list of functions *)
\end{Coqsrc}

@@

  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn: forall x y,  Ack x y <= Ack n (max x  y).

    Remark R01: n <> 0.

@@

    Remark R03: max 2 n = n.

    Remark R04: Ack n (2 * n + 3) = Ack n (Ack 2 n).

    Remark R05: Ack n (2 * n + 3) <= Ack (n + 2) n.

    Remark R06: Ack (n + 2) n <= Ack n (n + 2).

    Remark R07: Ack n (n + 2) < Ack n (2 * n + 3).
